---
title: "DOP : RNAseq Pipeline and Analysis"
output: html_notebook
---

```{r}
library(tximport)
library(readr)
library(DESeq2)
library(GenomicFeatures)
library(AnnotationDbi)
```

Upload RNA sample vs Time Condition
```{r}
setwd("/Users/hariniramanan/Desktop/DOP/RNAseq dataset")
samples <- read.csv("samples.csv", header=TRUE) 
```

Upload Quant files from Salmon Pseudo alignment
```{r}
files <- file.path("quants", samples$sample, "quant.sf")
names(files) <- samples$sample

```

Build a TxDb object from GTF -- getting transcript data from public

```{r}
txdb <- makeTxDbFromGFF("Homo_sapiens.GRCh38.110.gtf.gz")
```


Extract transcript-to-gene mapping
```{r}
k <- keys(txdb, keytype = "TXNAME")
tx2gene <- select(txdb, keys = k, keytype = "TXNAME", columns = "GENEID")
write.csv(tx2gene, file="tx2gene.csv", row.names=FALSE)
tx2gene <- read.csv("tx2gene.csv")
```
Now txi$counts will be gene-wise counts.
```{r}
txdb <- GenomicFeatures::makeTxDbFromGFF("Homo_sapiens.GRCh38.110.gtf.gz")
library(txdbmaker)
txdb <- txdbmaker::makeTxDbFromGFF("Homo_sapiens.GRCh38.110.gtf.gz")
```
Extract mapping with annotation file - create table with transcript and to its according gene
```{r}
k <- keys(txdb, keytype = "TXNAME")
tx2gene <- AnnotationDbi::select(txdb,
                                 keys = k,
                                 keytype = "TXNAME",
                                 columns = "GENEID")
```

Clean transcript IDs (remove version numbers like .1, .2)
```{r}
tx2gene$TXNAME <- sub("\\..*", "", tx2gene$TXNAME)
head(tx2gene)
write.csv(tx2gene, "tx2gene.csv", row.names = FALSE)
txi <- tximport(files, type = "salmon", tx2gene = tx2gene, ignoreTxVersion = TRUE)
```


```{r}
samples <- data.frame(
  SampleID = c("RMC192","RMC196","RMC200",
               "RMC193","RMC197","RMC201",
               "RMC194","RMC198","RMC202",
               "RMC195","RMC199","RMC203"),
  Timepoint = c("3hr","3hr","3hr",
                "3day","3day","3day",
                "6day","6day","6day",
                "10day","10day","10day"),
  stringsAsFactors = FALSE
)

dds <- DESeqDataSetFromTximport(
  txi,
  colData = samples,
  design = ~ Timepoint
)

dds <- DESeq(dds)


```

Gene Annotation
```{r}
library(ggplot2)
library(org.Hs.eg.db)
library(clusterProfiler)
library(AnnotationDbi)
library(UniProt.ws)
```



```{r}

res_3hr_vs_3day <- results(dds, contrast = c("Timepoint", "3hr", "3day"))
head(res_3hr_vs_3day)
res_3day_vs_6day <- results(dds, contrast = c("Timepoint", "3day", "6day"))
res_6day_vs_10day <- results(dds, contrast = c("Timepoint", "6day", "10day"))

# Define the contrasts you want to check
contrasts <- list(
  c("Timepoint", "3hr", "3day"),
  c("Timepoint", "3day", "6day"),
  c("Timepoint", "6day", "10day"),
  c("Timepoint", "3hr", "10day"),
  c("Timepoint", "3hr", "6day"),
  c("Timepoint", "3day", "10day")
)

```



```{r}
###Data Visualization 
library(ggplot2)  # nice package for volcano plots

# Define the contrasts
contrasts <- list(
  c("Timepoint", "3hr", "3day"),
  c("Timepoint", "3day", "6day"),
  c("Timepoint", "6day", "10day"),
  c("Timepoint", "3hr", "10day"),
  c("Timepoint", "3hr", "6day"),
  c("Timepoint", "3day", "10day")
)

library(ggplot2)

```



```{r}
  # 3hr vs 3day results
  res <- results(dds, contrast = c("Timepoint", "3hr", "3day"))
  res_df <- as.data.frame(res)
  
  # Add a column for significance (padj < 0.05 & |log2FC| > 1)
  res_df$Significant <- ifelse(res_df$padj < 0.05 & abs(res_df$log2FoldChange) > 1, "Yes", "No")
  
  # Volcano plot
  ggplot(res_df, aes(x = log2FoldChange, y = -log10(padj), color = Significant)) +
    geom_point(alpha = 0.6, size = 1.8) +
    scale_color_manual(values = c("No" = "grey", "Yes" = "red")) +
    theme_minimal() +
    labs(title = "Volcano Plot: 3hr vs 3day",
         x = "Log2 Fold Change",
         y = "-log10 adjusted p-value")
  
```



```{r}
  ## --- MA Plot ---
# MA plot
comp_name <- "3hr vs 3day"
plotMA(res, main = paste("MA Plot:", comp_name), ylim = c(-5, 5))

# PCA
vsd <- vst(dds)
plotPCA(vsd, intgroup = "Timepoint")

```



```{r}
library(pheatmap)
sampleDists <- dist(t(assay(vsd)))
sampleDistMatrix <- as.matrix(sampleDists)
rownames(sampleDistMatrix) <- colnames(vsd)
colnames(sampleDistMatrix) <- colnames(vsd)
pheatmap(sampleDistMatrix, clustering_distance_rows=sampleDists,
         clustering_distance_cols=sampleDists,
         main="Sample Distance Heatmap")
```

```{r}

topGenes <- head(order(res_3hr_vs_3day$padj), 50)
mat <- assay(vsd)[topGenes, ]
pheatmap(mat, scale="row", show_rownames=FALSE,
         annotation_col = as.data.frame(colData(dds)["Timepoint"]))

```
```{r}
# Gene Annotation
library(org.Hs.eg.db)
# Extract Ensembl IDs from results
ens_ids <- rownames(res_3hr_vs_3day)

# Map to SYMBOLS and ENTREZ IDs
gene_map <- mapIds(org.Hs.eg.db,
                   keys = ens_ids,
                   column = c("SYMBOL"),
                   keytype = "ENSEMBL",
                   multiVals = "first")

##GO ANALYSIS
library(clusterProfiler)
library(org.Hs.eg.db)

# Example: 3hr vs 3day results
res <- results(dds, contrast = c("Timepoint", "3hr", "3day"))
res_df <- as.data.frame(res)

# Filter significant DEGs
sig_res <- subset(res_df, padj < 0.05)

# Map Ensembl → Entrez
entrez_ids <- mapIds(org.Hs.eg.db,
                     keys = rownames(sig_res),
                     column = "ENTREZID",
                     keytype = "ENSEMBL",
                     multiVals = "first")

# Remove NAs
entrez_ids <- na.omit(entrez_ids)


ego <- enrichGO(gene          = entrez_ids,
                OrgDb         = org.Hs.eg.db,
                keyType       = "ENTREZID",
                ont           = "BP",      # Biological Process
                pAdjustMethod = "BH",
                pvalueCutoff  = 0.05,
                qvalueCutoff  = 0.05)

# View top results
head(ego)

```

```{r}
# Plot
barplot(ego, showCategory=20, title="GO Biological Process")
dotplot(ego, showCategory=20, title="GO Enrichment")

#KEGG
ekegg <- enrichKEGG(gene         = entrez_ids,
                    organism     = "hsa",   # hsa = Homo sapiens
                    pvalueCutoff = 0.05)
```


```{r}

barplot(ego, showCategory=20, title="GO Biological Process")

```

```{r}
##Venn Diagram
# Get DEGs for each timepoint vs 3hr
DEG_3day <- subset(results(dds, contrast=c("Timepoint","3day","3hr")), padj < 0.05)
DEG_6day <- subset(results(dds, contrast=c("Timepoint","6day","3hr")), padj < 0.05)
DEG_10day <- subset(results(dds, contrast=c("Timepoint","10day","3hr")), padj < 0.05)

# Extract gene IDs
genes_3hr  <- rownames(DEG_3day)  # DEGs appear here when 3day != 3hr
genes_3day <- rownames(DEG_3day)
genes_6day <- rownames(DEG_6day)
genes_10day <- rownames(DEG_10day)

library(VennDiagram)

venn.plot <- venn.diagram(
  x = list(
    "3hr"  = genes_3hr,
    "3day" = genes_3day,
    "6day" = genes_6day,
    "10day"= genes_10day
  ),
  filename = NULL,
  fill = c("red", "blue", "green", "purple"),
  alpha = 0.5,
  cex = 1.5,
  cat.cex = 1.5,
  main = "Overlap of DEGs across Timepoints"
)

grid::grid.draw(venn.plot)

```

